EGGs are adhesive!

### 2025-08-11 jlamp/jlamp.tex
minori

\author{Roberto Biondo}
\ead{r.biondo@studenti.unipi.it}

\author{Davide Castelnovo\corref{cor1}%
}
\ead{castelnovod@gmail.com}
%{https://orcid.org/0000-0002-5926-5615}{}

\author{Fabio Gadducci}
\ead{fabio.gadducci@unipi.it}
%{https://orcid.org/0000-0003-0690-3051}{}

@@

\affiliation{organization = {Department of Computer Science, University of Pisa},
	city = {Pisa}, 
	country = {Italy}}

\fntext[miur]{Research partially supported by the Italian MUR project PRIN
20228KXFN2 ``STENDHAL''.}

@@

\noindent
\begin{minipage}[l]{.6\linewidth}
	\begin{restatable}{lem}{mps}\label{prop:kerset}
		Suppose that in $\Set$ the commuting cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
		Then the following hold
		\begin{enumerate}
			\item the right face of the cube is a pullback;
			\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
		\end{enumerate}
	\end{restatable}
\end{minipage}  \hfill 
\begin{minipage}[r]{.35\linewidth}
	\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }
\end{minipage}

@@

\begin{prop}\label[prop]{prop:fact}
	Let $X$ be a set and $n\in \mathbb{N}$. Then the following facts hold
	\begin{enumerate}
		\item there are arrows $v_{n}\colon X^n\to X^\star $ such that $(X^\star, \{v_{n}\}_{n\in \mathbb{N}})$ is a coproduct;
		\item for every arrow $f\colon X\to Y$, $f^\star\colon X^\star \to Y^\star$ is the coproduct of the family $\{f^n\}_{n\in \mathbb{N}}$;
		\item $(-)^\star$ preserves all \emph{connected limits} \cite{carboni1995connected}, in particular it preserves pullbacks and equalizers.
	\end{enumerate}
\end{prop}

@@

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label[remark]{rem:length} It is well known that $(\mathbb{N}), \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathcal{N}\to 1^\star$. In particular, every set $X$ the arrow $!_\X\colon X\to 1$ induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside.
	\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=10pt{1 \ar[r]^{\id{1}} \ar[d]^{u_n}& 1 \ar[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=10pt@C=22pt{\mathbb{N} \ar[r]^{\Delta} \ar@{.>}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

@@

\begin{prop}[prop]\label{prop:com}
	$\hyp$ is isomorphic to $\comma{\id{\Set}}{\pro^\star}$
\end{prop}

@@

\begin{rem}\label{rem:functor}
	It is worth to point out, as first noted in \cite{bonchi2022string}, that $\hyp$ is equivalent to a category of presheaves. 
	%
	Indeed, consider the category $\catname{H}$ in which the set of objects is given by $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$ and arrows are given by the identities $\id{k,l}$, $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$. 
	%
	The functors $\catname{H}\to \Set$ corresponds exactly to hypergraphs: nodes correspond to the image of $\bullet$ while the set of hyperedges with source of length $k$ and target of length $l$ corresponds to the image of $(k,l)$ (see \cite{CastelnovoGM24}).
\end{rem}

In particular, \Cref{rem:functor} entails the following result.

@@

\begin{rem}It is worth mentioning that limits in $\hyp$ are not computed componentwise. By \Cref{prop:fact,prop:com,lim} we know that the two forgetful functors $U_{\hyp}, H_{\hyp}\colon \hyp \to \Set$, sending an hypergraph to, respectively, its set of nodes and that of hyperedges only jointly creates conneted limits. For instance, let $\phi\colon \mathbb{N}\to 1^\star$  $(\mathbb{N}, 1, \phi)$
\end{rem}

\iffalse
\subsubsection{$\hyp$ as a category of functors}

Following \cite{bonchi2022string}, we can present $\hyp$ as a category of functor over a suitable category.

\begin{defi}Let $\catname{H}$ be the category such that
	\begin{itemize}
		\item the set of objects is $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$;
		\item arrows are given by the identities $\id{k,l}$ and $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$;
		\item composition is defined by putting
		%\begin{equation*}
			$f_i=f_i\circ \id{k,l}$ and $f_i = \id{\bullet}\circ f_i$
		%\end{equation*}
		for every $f_i\colon (k,l)\rightarrow \bullet$.
	\end{itemize}
\end{defi}

The idea is that for every functor $F\colon \catname{H}\to \catname{Set}$ we can define
\[E_F:=\sum_{k,l\in \mathbb{N}}F(k,l)\]
%
Now, for every $k$, $l$, $i$ and $j$ in $\mathbb{N}$ with $i< k$ and $j< l$ we define $s^F_{k,l}\colon F(k,l)\to F(\bullet)^k$ and  $t^F_{k,l}\colon F(k,l)\to F(\bullet)^l$ as the unique arrows fitting in the diagrams below, where the vertical arrows are the projections
\[\xymatrix{F(k,l)  \ar@{.>}[r]^{s^F_{k,l}} \ar[dr]_{F(f_i)}& F(\bullet)^{k} \ar[d]^{\pi^F_{k,i}} & F(k,l) \ar@{.>}[r]^{t^F_{k,l}} \ar[dr]_{F(f_{k+j})} & F(\bullet)^{l} \ar[d]^{\pi^F_{l,j}} \\ & F(\bullet) && F(\bullet)}\]
%
In turn, these arrows allow us to consider
$s_F, t_F\colon E_F\rightrightarrows F(\bullet)^{\star}$ as the unique arrows fitting in the diagrams below, where the vertical arrows are coprojections
\[\xymatrix{F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{s^{F}_{k,l}}& F(\bullet)^{k} \ar[d]^{b^F_{k}} & F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{t^{F}_{k,l}}& F(\bullet)^{l} \ar[d]^{b^F_{l}}\\ E_F \ar@{.>}[r]_-{s_F}& F(\bullet)^\star & E_F \ar@{.>}[r]_-{t_F}& F(\bullet)^\star}\]

Let $\mathcal{G}_F$ be the resulting hypergraph. One can now show that  sending $F$ to $\mathcal{G}_F$ can be extended to an equivalence $\mathcal{G}_{-}\colon \Set^{\catname{H}}\to \hyp$ (see \cite{castelnovo2023thesis,CastelnovoGM24} for details).

\begin{prop}
	$\hyp$ is equivalent to the category $\Set^{\catname{H}}$.
\end{prop}
\fi

@@

\begin{defi} Given a set $X$, we say that two words $X^\star$ are \emph{disjoint} is they do not share any letter. An hypergraph $\mathcal{G}=(E_{\mathcal{G}}, V_{\mathcal{G}}, s_{\mathcal{G}}, t_{\mathcal{G}})$ is \emph{separated} if:
	\begin{enumerate}
		\item for every hyperedge $e$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ do not have repeated letters;
		\item for every two, not necessarily distinct, hyperedges $e, e'\in E_{\mathcal{G}}$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint;
		\item  for every two different hyperedges $e, e'\in E_{\mathcal{G}}$, $s_{\mathcal{G}}(e)$ and $s_{\mathcal{G}}(e')$ are disjoint;
		\item  for every two different hyperedges $e, e'\in E_{\mathcal{G}}$, $t_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint.
	\end{enumerate}
We will denote by $\shyp$ the full subcategory of $\hyp$ given by separated hypergraphs and by $S$ the inclusion functor.
\end{defi}
\iffalse 
\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.

 Moreover, let $\is{(V_{\mathcal{G}})}$ be the set of isolated nodes  of $\mathcal{G}$ i.e.~those which are not in the image of $s_{\mathcal{G}}$ or of $t_{\mathcal{G}}$. Furthermore, let $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}} \cup \{\iota_{is}\})$ be the coproduct of the family $\{V_{e}\}_{e\in E_\mathcal{G}} \cup \{\is{(V_{\mathcal{G}})}\}$ and define:
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of $1$.

@@

@@

				\draw[style=diredge, in=30, out=0] (timesout.center) to (v);
				\draw[style=diredge, in=180, out=0] (divout.center) to (d);
			\end{pgfonlayer}
		\end{tikzpicture}
	\end{center}

\begin{prop}There exists a functor $\sp \colon \hyp\to \shyp$ sending $\mathcal{G}$ to $\sp(\mathcal{G})$.
\end{prop}
\begin{proof}
Let $(h_1, h_2)\colon \mathcal{G}\to \mathcal{H}$ be a morphism of hypergraph. For every $a$

Preservation of compositions and identities follows at once from the construction.
Now, for every $e\in \mathcal{E}_\mathcal{G}$ and $i$ between $1$ and $n_e+m_e$  we can define 
\begin{align*}f_{e, i}\colon 1&\to V_{\mathcal{G}}\\
	* & \mapsto\begin{cases}
		(s_{\mathcal{G}}(e))_i & i \leq n_e\\
		(t_{\mathcal{G}}(e))_i & n_e < i
	\end{cases}
\end{align*}
These arrows, in turn, give us $f_ e\colon V_e\to V_{\mathcal{G}} $ and so an arrow $f_{\mathcal{G}}:V_{\sp(\mathcal{G})}\to V_{\mathcal{G}}$ such that 
\[f_{\mathcal{G}}\circ \iota_e = f_e \qquad f_{\mathcal{G}}\circ \is{\iota} =  j_{\mathcal{G}}\]
where $j_{\mathcal{G}}$ is the inclusion of $\is{(V_{\mathcal{G}})}\mto V_{\mathcal{G}}$. Notice that, for every $e\in E_{\mathcal{G}}$, $1\leq i\leq n_e$ and $1\leq k\leq m_e$ we have:
\[\begin{split}
	(f_{\mathcal{G}}^\star( s_{\sp(\mathcal{G})}(e)) )_i & = f_{\mathcal{G}}( s_{\sp(\mathcal{G})}(e) _i)\\
	&=f(\iota_e(\iota^e_i(*)))\\&=f_e(\iota^e_i(*))\\&=f_{e, i}(*)\\&=
	(s_{\mathcal{G}}(e))_i 
\end{split} \qquad 
\begin{split}
	(f^\star( t_{\sp(\mathcal{G})}(e)) )_k & = f( t_{\sp(\mathcal{G})}(e) _k)\\
	&=f(\iota_e(\iota^e_{n_e+k}(*)))\\&=f_e(\iota^e_{n_e+k}(*))\\&=f_{e, n_e+k}(*)\\&=
	(t_{\mathcal{G}}(e))_k
\end{split}
\]
and so we get an arrow $(\id{E_{\mathcal{G}}}, f_{\mathcal{G}})\colon V\to \mathcal{V}_{\mathcal{G}}$.

Now, let $\mathcal{H}=(E_{\mathcal{H}}, V_{\mathcal{H}}, s_{\mathcal{H}}, t_{\mathcal{H}})$ be another separated hypergraph with an arrow $(h_1, h_2)\colon S(\mathcal{H})\to \mathcal{G}$.  We can define
\begin{align*}
\hat{h}\colon V_{\mathcal{H}}&\to V_{\sp(\mathcal{G})} \\
v & \mapsto \begin{cases}
	\iota_{is}(h_2(v)) & h_2(v)\in \is{(V_{\mathcal{G}})}\\
	\iota_e(h_2(v)) & h_2(v) \text{ is a letter of } s_{\mathcal{G}}(e) \text{ for some } e\in E_{\mathcal{G}}\\
		\iota_e(h_2(v)) & h_2(v) \text{ is a letter of } t_{\mathcal{G}}(e) \text{ for some } e\in E_{\mathcal{G}}\\
\end{cases}
\end{align*}

We first have to show that the previous definition gives us a well-defined function.  By definition of isolated node we know that the three cases are exhaustive. Now, suppose that there are $e_1$ and $e_2$, not necessarily distinct, in $E_\mathcal{G}$ such that $h_2(v)$ is a letter of both

\noindent
\begin{minipage}[l]{.75\linewidth}
\setlength{\parindent}{1.5em}
\indent 	Since $\mathcal{H}$ is a separated hypergraph, $\hat{h}$
	is a well-defined function. Moreover,  by construction $(h_1, \hat{h})$ defines a morphism $\mathcal{H}\to V_{\sp(\mathcal{G})}$ fitting in the diagram aside.
\end{minipage} \hfill 
\begin{minipage}[r]{.25\linewidth}
	\xymatrix@C=40pt@R=10pt{S(\mathcal{H}) \ar[dr]^{(h_1, h_2)} \ar[d]_{S(h_1, \hat{h})}\\ S(\sp(\mathcal{G}))  \ar[r]_-{(\id{E_{\mathcal{G}}}, f_{\mathcal{G}})}& \mathcal{G}}
\end{minipage}

To prove uniqueness of such an arrow, let $(k_1, k_2)$ be a morphism $(\mathcal{H})\to \sp(\mathcal{G})$ such that 
\[(h_1, h_2)=(\id{E_{\mathcal{G}}}, f)\circ (k_1, k_2)\]
Clearly this implies $k_1 = h_2$.   Let now $v$ be an element of $V_\mathcal{H}$. Then we have three cases.
\begin{itemize}
	\item $v\in \is{(V_{\mathcal{H}})}$. In this case we have:
\begin{align*}
	j(h_2(v))&=f(\iota_{is}(h_2(v)))\\&=f(\hat{h}(v))\\&=h_2(v)\\&=
\end{align*}
	\item 
\end{itemize}

The previous three points allow us to conclude.
\end{proof}

@@

\fi

The following result will be useful. A proof of it is \Cref{proof:seplim}

@@

\subsection{Labelled hypergraphs}\label{sssect:hyp_alg_sign}

>>>>

\author[1]{Roberto Biondo}
\ead{r.biondo@studenti.unipi.it}

\author[1]{Davide Castelnovo\corref{cor1}%
}
\ead{castelnovod@gmail.com}
%{https://orcid.org/0000-0002-5926-5615}{}

\author[1]{Fabio Gadducci}
\ead{fabio.gadducci@unipi.it}
%{https://orcid.org/0000-0003-0690-3051}{}

@@

\affiliation[1]{organization = {Department of Computer Science, University of Pisa},
	addressline = {Largo B. Pontecorvo, 3},
	postcode = {56127},
	city = {Pisa}, 
	country = {Italy}}

@@

\noindent
\begin{minipage}[l]{.6\linewidth}
	\begin{restatable}{lem}{mps}\label{prop:kerset}
		Suppose that in $\Set$ the commuting cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
		Then the following hold true:
		\begin{enumerate}
			\item the right face of the cube is a pullback;
			\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
		\end{enumerate}
	\end{restatable}
\end{minipage}  \hfill 
\begin{minipage}[r]{.35\linewidth}
	\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }
\end{minipage}

@@

\begin{prop}\label[prop]{prop:fact}
	Let $X$ be a set and $n\in \mathbb{N}$. Then the following facts hold
	\begin{enumerate}
		\item there are arrows $v_{n}\colon X^n\mto X^\star $ such that $(X^\star, \{v_{n}\}_{n\in \mathbb{N}})$ is a coproduct;
		\item for every arrow $f\colon X\to Y$, $f^\star\colon X^\star \to Y^\star$ is the coproduct of the family $\{f^n\}_{n\in \mathbb{N}}$;
		\item $(-)^\star$ preserves all \emph{connected limits} \cite{carboni1995connected}, in particular it preserves pullbacks and equalizers.
	\end{enumerate}
\end{prop}

@@

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label{rem:length} It is well known that $(\mathbb{N}), \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathbb{N}\to 1^\star$. In particular, for every set $X$ the unique arrow $!_\X\colon X\to 1$ into the terminal induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside. 
	\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=8pt@C=30pt{1 \ar[r]^{\id{1}} \ar@{>->}[d]^{\delta_n}& 1 \ar@{>->}[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=8pt@C=28pt{\mathbb{N} \ar[r]^{\Delta} \ar@{<.}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

\begin{rem}\label{rem:le} Consider a function $f\colon X\to Y$  and suppose that $(X^\star, \{x_n\}_{n\in \mathbb{N}})$ and $\{Y^\star, \{y_n\}_{n\in \mathbb{N}}\}$ are the coproducts given by the first point of \Cref{prop:fact}, then for every $n\in \mathbb{N}$ we have 
	\begin{align*}
&\lgh_{Y}\circ f^\star \circ x_n = \lgh_{Y}\circ y_n \circ f^n= \Delta^{-1}\circ !^\star_{Y} \circ y_n\circ f^n =  \Delta^{-1}\circ v_n \circ !^n_{Y} \circ f^n \\=& \Delta^{-1}\circ v_n \circ (!_{Y} \circ f)^n  = \Delta^{-1}\circ v_n \circ (!_{X})^n = \Delta^{-1}\circ !^\star_{X} \circ x_n = \lgh_{X}\circ x_n
	\end{align*}
and so $\lgh_{Y}\circ f^\star= \lgh_{X}$.
\end{rem}

\noindent
\begin{minipage}[r]{.83\linewidth}
	\begin{rem}\label{rem:len} Let $X$ be a set.  It is worth noticing that, for every $n\in \mathbb{N}$, $X^n$ is the preimage of $n$ along $\lgh_{X}$. This amount to say that the square aside is a pullback. This follows at once from the fact that $\Set$ is \emph{extensive} (see \cite{carboni1993introduction} and also \Cref{word:ext}). 
	\end{rem}
\end{minipage}\hfill \begin{minipage}[l]{.15\linewidth}
\xymatrix@R=10pt{X^n \ar@{>->}[r]^{v_n} \ar[d]^{!_{X^n}}& X^\star  \ar[d]^{\lgh_{X}}\\ 1 \ar@{>->}[r]^{\delta_n} & \mathbb{N}}
\end{minipage}

@@

\begin{prop}\label{prop:com}
	$\hyp$ is isomorphic to $\comma{\id{\Set}}{\pro^\star}$
\end{prop}

@@

In order to deal with general limits of hypergraphs, it is worth to recall that, as first noted in \cite{bonchi2022string}, $\hyp$ is equivalent to a category of presheaves.

\begin{defi} We define the category $\catname{H}$  as the one in which the set of objects is given by $ (\mathbb{N}\times \mathbb{N}) + \{\bullet\}$ and arrows are given by the identities $\id{k,l}$, $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$. 
\end{defi}

It is possible to show that hypergraphs corresponds to presheaves over $\catname{H}$ (\cite{bonchi2022string,castelnovo2023thesis,CastelnovoGM24}). In \Cref{proof:equi} we report a proof of this claim for the interested reader.

\noindent 
\begin{minipage}[r]{.75\linewidth}
	\begin{restatable}{lem}{equi}\label{lem:equi}
		There is an equivalence of categories $\Set^{\catname{H}}\to \hyp$ sending a functor $F\colon \catname{H}\to \Set$ to $\mathcal{G}_F:=(E_F, F(\bullet), s_{F}, t_F)$, where $(E_F, \{\iota^F_{k,l}\})$ is a coproduct for $\{F(k,l)\}_{k,l\in\mathbb{N}}$
		and $s_F, t_F\colon E_F\rightrightarrows V^\star_F$ are the unique arrows fitting in the diagrams aside, where $(F(\bullet), \{v^F_n\}_{n\in \mathbb{N}})$ is the coproduct coming from the first point of\Cref{prop:fact} and $s^F_{k,l}, t^F_{k,l}\colon F(k,l)\to F(\bullet)^k$, $t^F_{k,l}\colon F(k,l)\to F(\bullet)^l$ are the arrows induced by, respectively, $\{F(f_i)\}_{i=0}^{k-1}$ and $\{F(f_i)\}_{i=k}^{k+l-1}$.
	\end{restatable}
\end{minipage}\hfill
\begin{minipage}[l]{.25\linewidth}
\xymatrix@R=10pt@C=40pt{F(k,l) \ar@{>->}[d]_{\iota^F_{k,l}}\ar[r]_{s^F_{k,l}} & F(\bullet)^k \ar@{>->}[d]^{v^F_k} \\ E_F \ar@{.>}[r]^{s^F} & F(\bullet)^\star}
\xymatrix@R=10pt@C=40pt{F(k,l) \ar[r]_{t^F_{k,l}} \ar@{>->}[d]_{\iota^F_{k,l}} & F(\bullet)^l\ar@{>->}[d]^{v^F_l}  \\ E_F \ar@{.>}[r]^{t^F} & F(\bullet)^\star}
\end{minipage}

In particular, \Cref{lem:equi} entails the following result.

@@

It is worth mentioning that limits in $\hyp$ which are not connected  are not computed componentwise. The following lemma shows how to compute products (see \Cref{proof:produ} for the proof).

	\begin{restatable}{lem}{produ}\label{lem:prod}
		Let $F\colon \D\to \hyp $ be a  discrete diagram. Let $F(D)$ be $(E_D, V_D, s_D,  t_D)$ and suppose that $(V, \{\pi^D_V\}_{D\in \D})$ is the product of $\{V_D\}_{D\in \D}$. For every $D\in \D$ and $(k,l)\in \mathbb{N}\times \mathbb{N}$, define
		\[E^D_{k,l}:=\{e\in E_D \mid \lgh_{V_D}(s_{D}(e)) = k \text{ and } \lgh_{V_D}(t_{D}(e)) = l \}\]
		with the canonical inclusion $i^D_{k,l}\colon E^D_{k,l}\mto E_D$.	Now, consider the product $(E_{k,l}, \{\pi^D_{k,l}\}_{D\in \D})$ and take the coproduct $(E_{\mathcal{L}}, \{\iota_{k,l}\}_{k,l \in \mathbb{N}})$ of the family $\{E_{k,l}\}_{k,l \in \mathbb{N}}$. Then there are arrows $s_{\mathcal{L}}, t_{\mathcal{L}}\colon E_{\mathcal{L}}\rightrightarrows V^\star$ and $\pi^D_E\colon E_{\mathcal{L}}\to E_D$ fitting in the diagram below and making $((E_{\mathcal{L}}, V, s_{\mathcal{L}}, t_{\mathcal{L}}), \{(\pi^D_E, \pi^D_V)\}_{D\in \D})$ a limit for $F$.
		\[\xymatrix@R=10pt@C=30pt{E^D_{k,l} \ar@{>->}[r]^{i^D_{k,l}} & E_D \ar[r]^{s_D} & V^\star_D  &E^D_{k,l} \ar@{>->}[r]^{i^D_{k,l}} & E_D \ar[r]^{t_D} & V^\star_D & E_{k,l} \ar[r]^{\pi^D_{k,l}} \ar[dd]_{\iota_{k,l}} & E^{D}_{k,l} \ar[dd]^{i^D_{k,l}}\\  E_{k,l} \ar@{>->}[d]_{\iota_{k,l}} \ar[u]^{\pi^D_{k,l}}  & & &E_{k,l} \ar@{>->}[d]_{\iota_{k,l}} \ar[u]^{\pi^D_{k,l}}  & & \\  E_{\mathcal{L}}   \ar@{.>}[rr]_{s_\mathcal{L}}& &V^\star \ar[uu]_{(\pi^D_V)^\star} &  E_{\mathcal{L}}   \ar@{.>}[rr]_{t_\mathcal{L}}& &V^\star \ar[uu]_{(\pi^D_V)^\star} & E_{\mathcal{L}} \ar@{.>}[r]_{\pi^D_E} & E_D}
		\]
	\end{restatable}

\begin{exa}\label{ex:term} For instance, a terminal object of $\hyp$ is given by $(\mathbb{N}\times \mathbb{N}, 1, \Delta\circ \pi_1, \Delta\circ \pi_2)$, where $\pi_1, \pi_2\colon \mathbb{N}\times \mathbb{N}\to \mathbb{N}$ are the projections and $\Delta\colon \mathbb{N}\to 1^\star$ the canonical isomorphism introduced in \Cref{rem:length}. 
\end{exa}

@@

\begin{defi} Given a set $X$, we say that two words $X^\star$ are \emph{disjoint} is they do not share any letter. An hypergraph $\mathcal{G}=(E_{\mathcal{G}}, V_{\mathcal{G}}, s_{\mathcal{G}}, t_{\mathcal{G}})$ is \emph{separated} if:
	\begin{enumerate}
		\item for every hyperedge $e$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ do not have repeated letters;
		\item for every two, not necessarily distinct, hyperedges $e, e'\in E_{\mathcal{G}}$, $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint;
		\item  for every two different hyperedges $e, e'\in E_{\mathcal{G}}$, $s_{\mathcal{G}}(e)$ and $s_{\mathcal{G}}(e')$ are disjoint;
		\item  for every two different hyperedges $e, e'\in E_{\mathcal{G}}$, $t_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e')$ are disjoint.
	\end{enumerate}
We will denote by $\shyp$ the full subcategory of $\hyp$ given by separated hypergraphs and by $S$ the inclusion functor.
\end{defi}

We begin by introducing a construction which ``split'' a given hypergraph to produce a separated one.

\begin{defi}
Given an hypergraph $\mathcal{\mathcal{G}}=(E_{\mathcal{G}}, V_{\mathcal{F}}, s_{\mathcal{G}}, t_{\mathcal{G}})$, for every hyperedge $e\in E_{\mathcal{G}}$ with $n_e=\lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ and $m_e=\lgh_{V_{\mathcal{G}}}(t_{\mathcal{G}}(e))$, we can consider the coproduct $(V_e, \{\iota^e_{i}\}_{i=1}^{n_e+m_e})$ of $n_e+m_e$ copies of $1$.

Let $(V_{\sp(\mathcal{G})}, \{\iota_e\}_{e\in E_{\mathcal{G}}} \cup \{\iota_{is}\})$ be the coproduct of the family $\{V_{e}\}_{e\in E_\mathcal{G}}$ and define:
\[\begin{split}
	s_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=1}^{n_e} \iota_e(\iota^e_{i}(0))
\end{split}\qquad \begin{split}
	t_{\sp(\mathcal{G})} \colon E_{\mathcal{G}}&\to V_{\sp(\mathcal{G})}^\star\\
	e &\mapsto \prod_{i=n_e+1}^{m_e} \iota_e(\iota^e_{i}(0))
\end{split} \]
where $0$ is the unique element of $1$.

@@

\begin{exa}\label{ex:empty}
	If $\mathcal{G}$ has no hyperedges, then $\sp(\mathcal{G})$ is the empty hypergraph.
\end{exa}

@@

				\draw[style=diredge, in=30, out=0] (timesout.center) to (v);
				\draw[style=diredge, in=180, out=0] (divout.center) to (d);
			\end{pgfonlayer}
		\end{tikzpicture}
	\end{center}
\end{exa}

This construction can be extended to a functor.

\begin{prop}There exists a functor $\sp \colon \hyp\to \shyp$ sending $\mathcal{G}$ to $\sp(\mathcal{G})$.
\end{prop}
\begin{proof}
Let $(h_1, h_2)\colon \mathcal{G}\to \mathcal{H}$ be a morphism of hypergraphs.  For every hyperedge $e\in E_{\mathcal{G}}$,
the universal property of coproducts, there are arrows $V_e\to V_{h_1(e)}$ fitting in the diagram below, on the left. These, in turn, yield the dotted $V_{\sp(\mathcal{G})} \to V_{\sp(\mathcal{G})}$ on the right.
\[\xymatrix{1 & 1 \\ V_{e} & V_{h_1(e)} }\]
Preservation of compositions and identities follows at once from the construction.
Now, for every $e\in \mathcal{E}_\mathcal{G}$ and $i$ between $1$ and $n_e+m_e$  we can define 
\begin{align*}f_{e, i}\colon 1&\to V_{\mathcal{G}}\\
	* & \mapsto\begin{cases}
		(s_{\mathcal{G}}(e))_i & i \leq n_e\\
		(t_{\mathcal{G}}(e))_i & n_e < i
	\end{cases}
\end{align*}
These arrows, in turn, give us $f_ e\colon V_e\to V_{\mathcal{G}} $ and so an arrow $f_{\mathcal{G}}:V_{\sp(\mathcal{G})}\to V_{\mathcal{G}}$ such that 
\[f_{\mathcal{G}}\circ \iota_e = f_e \qquad f_{\mathcal{G}}\circ \is{\iota} =  j_{\mathcal{G}}\]
where $j_{\mathcal{G}}$ is the inclusion of $\is{(V_{\mathcal{G}})}\mto V_{\mathcal{G}}$. Notice that, for every $e\in E_{\mathcal{G}}$, $1\leq i\leq n_e$ and $1\leq k\leq m_e$ we have:
\[\begin{split}
	(f_{\mathcal{G}}^\star( s_{\sp(\mathcal{G})}(e)) )_i & = f_{\mathcal{G}}( s_{\sp(\mathcal{G})}(e) _i)\\
	&=f(\iota_e(\iota^e_i(*)))\\&=f_e(\iota^e_i(*))\\&=f_{e, i}(*)\\&=
	(s_{\mathcal{G}}(e))_i 
\end{split} \qquad 
\begin{split}
	(f^\star( t_{\sp(\mathcal{G})}(e)) )_k & = f( t_{\sp(\mathcal{G})}(e) _k)\\
	&=f(\iota_e(\iota^e_{n_e+k}(*)))\\&=f_e(\iota^e_{n_e+k}(*))\\&=f_{e, n_e+k}(*)\\&=
	(t_{\mathcal{G}}(e))_k
\end{split}
\]
and so we get an arrow $(\id{E_{\mathcal{G}}}, f_{\mathcal{G}})\colon V\to \mathcal{V}_{\mathcal{G}}$. 
\end{proof}

@@

\begin{defi} 
	isolati e connessi
\end{defi}

\begin{prop}
	aggiunto dx
\end{prop}

Finally, we conclude our exam of separated hypergraph proving a result regarding connected limits. A proof of it is \Cref{proof:seplim}

@@

\begin{rem}\label{rem:map1}
	diagramma generale dei vari ipergrafi
\end{rem}
\subsection{Labelled hypergraphs}\label{sssect:hyp_alg_sign}
